Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-10 paper.tex
revisado todo artigo, falta apenas apendices

\section{Concluding remarks and future work}\label{conc}
\opsi, based in \vcps,
yields initial steps in achieving an effective social participation ontology: community has registered activities and delivered reference documents, including this article.
Sections~\ref{ospUtil},~\ref{downwards},~\ref{upper} and~\ref{public} have expansion directions for the ontology, regarding uses, further definition of classes, upper ontologies and related ontological structures.

On the practical side, the use of this ontology or related developments for the Brazilian federal participation portal (\participa~\cite{participa}) is a desirable reality, as it implies usage and good maintenance.  Moreover, an ontology was done for \url{Participa.br}, based in the \ops: the \opa~\cite{opa} (Ontology of \url{Participa.br}). This is confluent with the presidential Decree that establishes a policy and commitment for social participation\cite{decree}. In this context, presidential and ministerial parties have come close, to start formalizing current legal participatory mechanisms (e.g. conferences, councils, forums, public consultations, round tables) in ontological terms, which resulted in the Social Library Ontology (\obs) and the Social Library Vocabulary (\vbs).
Hosting of ontologies on Webprotege~\cite{webprotege} have become central, as a way to share specific ontologies in a friendly environment and to collects observations.

\subsection{Further work}
Further works involves observing community manifestations about \ops\ and this article, accomplishing use by means of formal instances and civil society, studying upper ontologies usage with \ops\. The use of \ops\ (or a variant) in different instances is being tackled for the creation of social participation indicators and easing participation processes.

Academic texts dedicated to each new participation ontology created (\ocd, \obs, \ontologiaa, \opa) should be written and submitted to peer review for enhancements and quality assurance.

Data related to these ontologies were found in relational databases and preliminary scripts were written to make them available as \rdf. A sound linkage of this data and relation to the Linked Open Data (\lod~\cite{}) cloud is planed for a near future. This should make Brazilian participative structures and data part of the Giant Global Graph~\cite{ggg}.

\begin{acknowledgments}
This article is deeply influenced by Protege and BFO documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Renato Fabbri is grateful to CNPq (process 140860/2013-4, project 870336/1997-5), PNUD/ONU (contract 2013/00056, project BRA/12/018), SNAS/SG-PR, and the Postgraduate Committee of the IFSC/USP. Authors thank \corais\ Platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation. Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.
% 2013/00056, projeto BRA/12/018 (pelo que?? nao me lembro mais e nao anotamos)
\end{acknowledgments}

>>>>

\section{Concluding remarks and future work}\label{conc}
\opsi, based in \vcps,
yields initial steps in achieving an effective social participation ontology: community has registered activities and delivered reference documents, including this article.
Sections~\ref{exp} and~\ref{ospUtil} have expansion directions for the ontology, regarding uses, further definition of classes, upper ontologies and related ontological structures.

On the practical side, the use of this ontology or related developments for the Brazilian federal participation portal (\participa~\cite{participa}) is a desirable reality, as it implies usage and good maintenance.  Moreover, an ontology was done for \participa, based in the \ops: the \opa~\cite{pnud5} (Ontology of \participa). This is confluent with the presidential Decree that establishes a policy and commitment for social participation\cite{decree}. In this context, presidential, ministerial and academic parties started formalizing current legal participatory mechanisms (e.g. conferences, councils, forums, public consultations, round tables) in ontological terms, which resulted in the Social Library Ontology (\obs\ from the Brazilian name Ontologia da Biblioteca Social) and the Social Library Vocabulary (\vbs\ from the Brazilian name Vocabul\'ario da Biblioteca Social).
Hosting ontologies on Webprotege~\cite{webprotege} have become central, as a way to share specific ontologies in a friendly environment and to collects observations.

\subsection{Further work}
Further work involves observing community manifestations about \ops\ and this article, accomplishing use by means of formal instances and civil society, and studying upper ontologies usage with \ops. The use of \ops\ (or a variant) in different instances is being tackled for the creation of social participation indicators and easing participation processes.

Academic texts dedicated to each new participation ontology created (\ocd, \obs, \ontologiaa, \opa) should be written and submitted to peer review for enhancements and quality assurance~\cite{pnud5}.

Data related to these ontologies were found in relational databases and preliminary scripts were written to make them available as \rdf. A sound linkage of this data and incorporation to the Linked Open Data (\lod~\cite{lod}) cloud is planed for a near future. This should make Brazilian participative structures and data part of the Giant Global Graph~\cite{ggg}.

\begin{acknowledgments}
This article is deeply influenced by Protege and BFO documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Authors thank \corais\ Platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation. Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.
Renato Fabbri is grateful to CNPq (process 140860/2013-4, project 870336/1997-5), PNUD/ONU (contract 2013/00056, project BRA/12/018), SNAS/SG-PR, and the Postgraduate Committee of the IFSC/USP. 
% 2013/00056, projeto BRA/12/018 (pelo que?? nao me lembro mais e nao anotamos)
\end{acknowledgments}
